---
title: "Test CLT with Exponentials"
author: "JeremiShane"
date: "11/24/2017"
output:
  pdf_document: default
  html_document: default
---

## Summary
In testing the Central Limit Theorem (CLT) with means from Exponential distributions, the evidence shows the CLT to hold true.  

## Overview
The central limit theorem (CLT) establishes that, in most situations, when independent random variables are added, their properly normalized sum tends toward a normal distribution (informally a "bell curve") even if the original variables themselves are not normally distributed.  We will evaluate this theory with the averages/means of simulated exponentials.  

From Wikipedia, the exponential distribution occurs naturally when describing the lengths of the inter-arrival times in a homogeneous Poisson process.  The exponential distribution describes the time for a continuous process to change state

This analysis is organized as follows:  
1. Simulations  
2. Sample Mean versus Theoretical Mean  
3. Sample Variance versus Theoretical Variance  
4. Distribution  

First, let's have a quick look at an exponential distribution and a normal distribition.

```{r expvsnormal, echo=FALSE}
set.seed(23); lambda <- 0.2  ## lambda is our rate
par(mfrow=c(1,2), oma=c(0,0,0,0))
randomexponential <- rexp(10000, lambda); re <- density(randomexponential)
plot(re, type="n", main="random exponential density")
polygon(re, col="red", border="grey")
abline(v=mean(randomexponential), col="black", lwd=2)
legend('topright', "mean", lty=1, col="black", lwd=2)
set.seed(23); randomnormal <- rnorm(10000); rn <-density(randomnormal)
plot(rn, type="n", main="random normal density")
polygon(rn, col="red", border="grey")
abline(v=mean(randomnormal), col="black", lwd=2)
```  

We can see from these two simple histograms that an exponential distribution does not look normal.  To carry out our experiment, we will look at the means of multiple exponential distributions.  The Central limit theorem tells us that even though the exponential distribution is not normal, the means from large enough number of exponential distributions should approach normal.     

## Simulation of Exponentials
Below is the R code for simulating the exponential distribution with size = 40 and 100.  In each case we perform 1000 simulations and take the mean of each simulation.  

```{r p1simulation}
set.seed(23)  ## set seed for reproducability
## run 1000 simulations
nsims <- 1000
n <- 40 ## our default number of exponentials
forty_mean = sapply(1:nsims, function(x) {mean(rexp(n, lambda))})
hundred_mean = sapply(1:nsims, function(x) {mean(rexp(100, lambda))})
```
  
## Simulated Mean vs. Theoretical Mean
With rate lambda = .2, an exponential distribution should have mean = 1/lambda = 5, and standard deviation = 1/lambda as well.  We would expect the means from 1000 of these exponential distributions to be a normal distribution around 5, from the Central Limit Theorem.  

```{r theMeans, echo=FALSE}
meansdf <- data.frame("theoreticalMean"=1/lambda)
meansdf$fortyMean <- mean(forty_mean)
meansdf
```  
These calculations show that the theoretical mean and the mean from each of our simulations are very close.

## Simulated Variance vs. Theoretical Variance
Variance...  
- of an exponential distribution is 1/lambda^2   
- for a normal distribution is sd^2  

```{r theVariance, echo=FALSE}
variancedf <- data.frame("theoreticalVariance" = 1/lambda^2/n)
## 1/lambda^2/n  ## our theoretical variance for 1000 means of n=40 size exponentials
variancedf$simulatedVariance <- var(forty_mean)  ## the actual variance of our 40 size exponential simulation
variancedf
```   
We see here that the theoretical variance and calculated variance from our simulated distribution is very close.    

## Distribution  
Let's look at a histogram distribution and a qq plot of our forty_mean simulation.  Both of these indicate the Central Limit Theorem is holding true.  The distribution is forming a normal bell curve around 5.

Then, let's refer back to our exponential distribution of size = 1000, and look at it alongside our distribution of the means of our simulations as n gets larger.  

The larger n gets, the closer to normal it gets.  It is safe to say that our experiment shows clearly the Central Limit Theorem holding true. 

```{r histogram, echo=FALSE}
par(mfrow=c(1,2), oma=c(0,0,0,0))
hist(forty_mean, breaks=10, prob=TRUE, main="Histogram") 
lines(density(forty_mean), col="black")
## the theoretical center of distribution is 1/lambda (vertical line at mean)
abline(v=1/lambda, col="red", lty=1, lwd=2)
## our simulated mean/center
abline(v=mean(forty_mean), col="black", lwd=1.5)
legend('topright', c("simulated mean", "theoretical mean"), lty=c(1,1), col=c("black", "red"))
qqnorm(forty_mean); qqline(forty_mean)
```  

***  

```{r distribution, echo = FALSE}
par(mfrow=c(1,3), oma=c(0,0,0,0))
randomexponential <- rexp(10000, lambda)
re <- density(randomexponential)
plot(re, type="n", main="random exponential density")
polygon(re, col="red", border="grey")
abline(v=mean(randomexponential), col="black", lwd=2)
legend('topright', "mean", lty=1, col="black", lwd=2)
d <- density(forty_mean)
plot(d, type="n", main="forty mean simulation")
polygon(d, col="red", border="grey")
abline(v=mean(forty_mean), col="black", lwd=2)
d <- density(hundred_mean)
plot(d, type="n", main="hundred mean simulation")
polygon(d, col="red", border="grey")
abline(v=mean(hundred_mean), col="black", lwd=2)
```   
